Tutorial_2a

### 2024-11-18 Tutorial_2a.Rmd
update html and Rmd files
update Tutorial 2a and ms.

---
title: "Tutorial_2a"
author: "Sven Panis"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

@@

In this Tutorial 2a we plot priors for the logit and cloglog link functions, build Bayesian hazard regression models for the first experiment of Panis and Schmidt (2016) using only the no-mask trials (prime = blank, congruent, or incongruent), and visualize the posterior distributions of the effects of congruent and incongruent prime types, relative to blank prime trials for the selected model.

Content:
1. Load libraries.
2. Load and wrangle the person-trial-bin data set that we saved in Tutorial 1a.
3. Plot the logit and complementary log-log (cloglog) link functions.
4. Visualize different prior distributions on the logit and cloglog scales.
5. Fit hazard models.
6. Compare models using loo and waic.
7. Display effects for the selected model.

@@

@@

```{r load-data}
ptb_data <- read_csv("Tutorial_1_descriptive_stats/data/inputfile_hazard_modeling.csv")
print(ptb_data,n=30)
summary(ptb_data) # 26602 rows: 6 participants, trial, 3 conditions, 15 periods, and event indicator (0/1)
```

Prepare the data set.

```{r prepare-data}
# select analysis time range: (200,600] with 10 bins (time bin ranks 6 to 15)
ptb_data <- ptb_data %>% filter(period > 5)

# create factor condition, with "blank" as the reference level
ptb_data <- ptb_data %>% mutate(condition = factor(condition, labels = c("blank", "congruent","incongruent")))

# center TIME (period) on bin 9, and trial on trial 1000 and rescale; Add dummy variables for each bin.
ptb_data <- ptb_data %>% 
        mutate(period_9 = period - 9,
               trial_c = (trial - 1000)/1000,
               d6  = if_else(period == 6, 1, 0),
               d7  = if_else(period == 7, 1, 0),
               d8  = if_else(period == 8, 1, 0),
               d9  = if_else(period == 9, 1, 0),
               d10 = if_else(period == 10, 1, 0),
               d11 = if_else(period == 11, 1, 0),
               d12 = if_else(period == 12, 1, 0),
               d13 = if_else(period == 13, 1, 0),
               d14 = if_else(period == 14, 1, 0),
               d15 = if_else(period == 15, 1, 0))

head(ptb_data,n=17)
summary(ptb_data)
# 6 subjects
# condition: blank (6401), congruent (2642), incongruent (3797)
# period_9: -3 to 6
# trial_c: -0.999 to 0.540
```

# 5. Fit hazard models.

## Model M1: general specification of TIME in the baseline condition (blank prime), and main effects of prime type, and trial number.

>>>>

---
title: "Tutorial_2a"
author: "Sven Panis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: united
editor_options: 
  chunk_output_type: console
---

@@

Tutorial 2a shows various ways to fit Bayesian binary regression models to discrete time-to-event data. The data are taken from the first experiment of Panis and Schmidt (2016), but using only the no-mask trials (prime = blank, congruent, or incongruent). We compare models using WAIC and LOO, and interpret parameter estimates for one model. 
We also plot the logit and cloglog link functions, and illustrate how various prior distributions on the logit and cloglog scales look on the original probability (i.e., hazard) scale.

@@

theme settings for ggplot

```{r plot-settings, results='hide'}
theme_set(
  theme_bw() +
    theme(text = element_text(size = 22, face = "bold"), 
          title = element_text(size = 22, face = "bold"),
          legend.position = "bottom")
)

## Set the amount of dodge in figures
pd <- position_dodge(0.7)
pd2 <- position_dodge(1)
```

@@

```{r load-data}
ptb_data <- read_csv("Tutorial_1_descriptive_stats/data/inputfile_hazard_modeling.csv")
print(ptb_data,n=30)
summary(ptb_data) # 26602 rows: 6 participants, trial (experiment-wise), 3 conditions, 15 periods, and event indicator (0/1)
```

Wrangle the data set.

```{r wrangle-data}
ptb_data <- ptb_data %>% 
# select analysis time range: (200,600] with 10 bins (time bin ranks 6 to 15)
  filter(period > 5) %>%

# create categorical predictor for TIME named "timebin" with index coding
  mutate(timebin = factor(period,levels=c(6:15)),
# create continuous predictor for TIME named "period_9", centered on bin 9,
         period_9 = period - 9,
# create binary variables "dummys" to indicate each bin in the analysis
         d6  = if_else(period == 6, 1, 0),
         d7  = if_else(period == 7, 1, 0),
         d8  = if_else(period == 8, 1, 0),
         d9  = if_else(period == 9, 1, 0),
         d10 = if_else(period == 10, 1, 0),
         d11 = if_else(period == 11, 1, 0),
         d12 = if_else(period == 12, 1, 0),
         d13 = if_else(period == 13, 1, 0),
         d14 = if_else(period == 14, 1, 0),
         d15 = if_else(period == 15, 1, 0),

# create continuous predictor for trial number named "trial_c", centered on bin 1000 and rescaled
         trial_c = (trial - 1000)/1000,
# create categorical predictor for trial number named "stage" (early,middle,late) with index coding
         stage = ifelse(trial <= 500, 1, ifelse(trial > 1000, 3, 2)),
         stage = factor(stage, levels=c(1,2,3)),

# create factor "condition", with "blank" as the reference level
         condition = factor(condition, labels = c("blank", "congruent","incongruent")),
# create categorical predictor "prime" with index-coding
         prime = ifelse(condition=="blank",1, ifelse(condition=="congruent",2,3)),
         prime = factor(prime,levels=c(1,2,3))) %>%
  select(pid,trial,trial_c,stage,condition,prime,period,period_9,timebin,d6:d15)

head(ptb_data,n=17)
summary(ptb_data)
```

# 3. Fit hazard models.

## Model M0: A null model without experimental predictors

The first model we fit is a "random intercepts" model, where we fit a single intercept for each timebin and allow each intercept to vary between participants.
This is a general specification of TIME, and can be used if we do not want to make assumptions about how cloglog-hazard changes over time (within a trial). 
There are two ways to implement such a model. First, we can use the index-coding approach, as .. .by McElreath...

Prepare the data file, and specify priors.

```{r data-priors-M0}
# remove unnecessary columns before fitting a model
M0_data <- ptb_data %>% select(pid, event, timebin)

priors_M0 <- c(
  set_prior("skew_normal(-0.59,1.26,-4.22)", class = "b"), # flat prior for p
  set_prior("normal(0, 1)", class = "sd"),                    
  set_prior("lkj(2)", class = "cor")                        
)
```

Fit the model (M0). old model M0d

```{r fit-model-M0, eval=F}
plan(multicore)

model_M0 <-
   brm(data = M0_data,
       family = binomial(link="cloglog"),
       event|trials(1) ~ 0 + timebin + (0 + timebin | pid),
       prior = priors_M0,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "models/model_M0")
```

2582.7 seconds.

Second, we can fit the same model using a dummy coding approach, as follows.

Prepare the data file, and specify priors.

Fit the model with dummy coding (M0d).

### Make assumptions to simplify the model

Instead of using a general specification of TIME, we can treat TIME as a continuous variable and make assumptions about how hazard changes over time within a trial. For example, we might assume that hazard changes in a linear + quadratic fashion over timebins within a trial.

Prepare the data file, and specify priors.

Fit the model with a continuous predictor (M0c).

## Model M1.

The next model we fit also includes our predictor variable prime-target congruency.
First, we can use index coding for the categorical predictor.

Second, if we do not want to use index coding, assume that hazard can vary in a linear+quadratic fashion over time, and relax the proportionality assumption, then we can fit the following model.

## Model M2.

The next model we fit also includes our predictor trial number.

First, we can categorise the continuous predictor trial number by grouping them in one of three "stages" of the experiment (stage 1 = trials 1 to 500; stage 2 = trials 501 to 1000; stage 3 = trials 1001 and later) and use index coding.

Second, we can treat trial number and bin number as continuous variables, and center them.

# Model comparisons.

# Model inference.
